---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r echo = TRUE, warning = FALSE, message = FALSE}
    library(readr)
    library(dplyr)
    library(ggplot2)
    library(lubridate)
```

## Loading and preprocessing the data
```{r echo = TRUE}
    unzip("activity.zip", overwrite = TRUE)
    activity <- read_csv("activity.csv") %>% 
        mutate(interval = sprintf("%04d", interval))
    activity$interval <- as.POSIXct(strptime(x = activity$interval, "%H%M"))
```


## What is mean total number of steps taken per day?

This prepares the data by summing the steps per day.
```{r echo = TRUE, warning = FALSE}
    
    activityTotals <- activity %>% 
        select(date,  steps) %>% 
        group_by(date) %>% 
        summarise(total_steps = sum(steps, na.rm = TRUE))
```

1. Make a histogram of the total number of steps each day
```{r echo = TRUE, warning = FALSE}
    qplot(
        x = activityTotals$date,
        xlab = "Date",
        y = activityTotals$total_steps, 
        ylab = "Total Steps",
        geom = "histogram",
        stat = "identity",
        fill = I("steelblue"))

```

2. Calculate and report the **mean** and **median** total number of steps each day.
```{r echo = TRUE, warning = FALSE}
    mean(activityTotals$total_steps)
    median(activityTotals$total_steps)
```

## What is the average daily activity pattern?

```{r echo = TRUE, warning = FALSE}
    
    activityByPeriod <- activity %>% 
        select(interval, steps) %>% 
        group_by(interval) %>% 
        summarise(mean_steps = mean(steps, na.rm = TRUE))
```

1. Make a time series plot of the average number of steps taken in each interval across all days
```{r echo = TRUE, warning = FALSE}
    qplot(data = activityByPeriod,
          x = interval,
          y = mean_steps,
          xlab = "Interval",
          ylab = "Average Steps",
          geom = "line")

```



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
